---
title: "Classical Decomposition Algorithm (CDA)"
output:
  word_document: default
  html_notebook: default
---

Classical decomposition algorithm (CDA) is performed upon a time series of monthly temperatures recorded in Columbus, OH from January 1995 to December 2005.
The units of measurement is Fahrenheit.

#here are the time series values for the temperature, 'temp'.

```{r}
temp <-
  c(29.9, 28.1, 44.3, 50.6, 61.0, 72.4, 75.6, 76.9, 63.6, 55.7, 38.1, 29.2,
    27.9, 30.7, 34.4, 50.5, 60.6, 70.8, 73.9, 73.6, 65.8, 55.5, 37.3, 37.4,
    28.0, 35.9, 42.9, 47.9, 57.1, 70.0, 74.3, 70.4, 65.4, 55.3, 40.6, 35.4,
    37.0, 39.4, 42.9, 53.7, 67.2, 71.7, 74.9, 75.7, 71.5, 56.0, 46.0, 41.5,
    31.3, 36.2, 36.9, 54.7, 64.8, 74.5, 79.2, 73.7, 68.3, 54.7, 48.6, 34.6,
    28.0, 36.5, 46.5, 50.5, 64.5, 71.7, 71.9, 71.3, 65.8, 57.5, 42.1, 24.4,
    27.6, 35.9, 37.1, 55.8, 63.3, 69.9, 74.5, 75.2, 65.3, 55.8, 49.2, 40.9,
    33.4, 38.1, 39.6, 55.0, 57.2, 73.0, 77.8, 76.2, 70.7, 55.0, 41.6, 31.8,
    24.6, 27.0, 43.0, 52.7, 61.1, 66.6, 73.2, 73.7, 65.7, 52.3, 49.0, 33.7,
    26.7, 29.1, 42.3, 52.1, 65.8, 69.7, 73.0, 70.8, 68.5, 55.0, 48.7, 32.4,
    32.7, 33.3, 34.9, 54.8, 57.4, 72.7, 76.8, 76.4, 71.0, 56.3, 45.7, 30.3)
```

## set up the time variable, 'year'

```{r}
year <- seq(from=1995, by=1/12, length=length(temp))
```

## plot the time series.

```{r}
plot(year, temp, type="l",
     xlab="year", ylab="monthly temperature (Fahrenheit)")
```

Part (a)
Summarize the plot of the time series -- what features of the data are present?

The plot of monthly temperatures vs year of Columbus recorded from 1995 to 2005 shows,
i. a unifrom trend with mean zero
ii. the variance is uniform
iii. there is a seasonality component during each year
iv. there are no outliers
v. there are no abrupt changes in behaviour

We will now carry out the classical decomposition algorithm.
** Step 1 of the algorithm **
Estimate the trend by smoothing the time series.

## let us choose the period to be d=12 months

```{r}
d <- 12
```

## half.d is d/2 
```{r}
half.d <- d/2
```

## smooth the temp time series using the 'filter' command
```{r}
our.filter <- c(0.5, rep(1,d-1), 0.5)/d
```

## display 'our.filter' to see the values makes sense!
```{r}
our.filter
```

## 'n' is the length of time series
```{r}
n <- length(temp)
```

## We will not filter the variable 'temp',
## but a new variable called 'longer.temp'
```{r}
longer.temp <- c(temp[(d/2):1], temp, temp[n:(n-d/2+1)])
longer.temp
```

## now filter 'longer.temp'.
```{r}
temp.MA <- filter(longer.temp, our.filter)[(d/2+1):(d/2+n)]
temp.MA
```

Part (b)
Try to explain why we have the variable 'longer.temp'. What does this variable contain, and what does
the variable allow us to do in the 'filter' command?

The variable 'longer.temp' contains all the values of temperatures from the original 'temp' time series plus the first 6 and last 6 values from the same 'temp' variable. So the 'longer.temp' varaiable contains 12 more values than the 'temp' variable. The 'longer.temp' variable is created because the filtering goes 6 values before the first value and 6 values after the 'temp' variable. In order to compensate for the filter, 'longer.temp' variable is created.

## plot the time series
```{r}
plot(year, temp, type="l",
     xlab="year", ylab="monthly temperature (Fahrenheit)")
```
##overlay the smooth
```{r}
plot(year, temp, type="l",
     xlab="year", ylab="monthly temperature (Fahrenheit)")
lines(year, temp.MA, type="l", lwd=2)
```
Part (c)
## Summarise the plot of the smooth -- what do you see?  Do you think the variance of the trend estimate is constant with time?  Explain.
The smoothed time series plot is obtained after removing the trend and variance of the trend estimate is constant with time.

** Step 2 of the algorithm **
## Estimate the seasonal component.

## calculate and plot the residuals of the smooth.
```{r}
temp.MA.resids <- temp - temp.MA
```

## plot the time series (first plot)
```{r}
plot(year, temp.MA.resids, type="l",
     xlab="year", ylab="residuals of the smooth")
```
## we define a variable to hold the names of the months
```{r}
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
            "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
```

## Now we need to estimate the average seasonal components.
## There should be 'd=12' of them. We first form 'temp.MA.resids' into a matrix with 'd' rows.
```{r}
temp.matrix <- matrix(temp.MA.resids, nrow=d)
temp.matrix
```
## create the plot of temperatures by month (second plot) in a panel of size 3 rows by 4 columns.
```{r}
par(mfrow=c(3,4)) 
for (m in 1:12) {
  plot(1995:2005, temp.matrix[m,], type="l",
       xlab="year", ylab="temperature", main=months[m])
  abline(h=mean(temp.matrix[m,]), lty=2)
}
```

Part (d)
## Describe the two plots above. Do we have evidence of seasonality in the time series?  Explain.
The first plot shows the residuals of the time series plot after removing the trend to check for seasonality and the second plot is residuals of the temperatures of different months plotted against the years between 1995-2005 and is same as the first plot obtained after removing the trend and to check for seasonality.From the second plot it can be seen that there is seasonality in the time series.

## now average across the years to get the 's.hat.stars'
```{r}
s.hat.stars <- rowMeans(temp.matrix)
s.hat.stars
```
## plot the 's.hat.stars', without the x-axis
```{r}
par(mfrow=c(1,1))
plot(s.hat.stars, type="l", xaxt="n")
```
## Now  we'll label the x-axis
```{r}
par(mfrow=c(1,1))
plot(s.hat.stars, type="l", xaxt="n")
axis(side=1, at=1:12, labels=months)
```
## now subtract off the mean of the 's.hat.stars' so that we have a
## seasonal component that adds up to zero (has average zero).
```{r}
s.hat <- s.hat.stars - mean(s.hat.stars)
```

## plot the 's.hat', without the x-axis
```{r}
plot(s.hat, type="l", xaxt="n",
     xlab="month", ylab="estimate of seasonal component")
```
## now we'll label the x-axis
```{r}
plot(s.hat, type="l", xaxt="n",
     xlab="month", ylab="estimate of seasonal component")
axis(side=1, at=1:12, labels=months)
```
## Part (e)
## Describe the estimate of the seasonal component in words.
The estimate of the seasonal componenet is obtained by subtracting the mean of the average value of residuals over all years from the average values of residuals across all years.
From the plot, it can be seen that the seasonality rises from January to July and descends after July.

## ** Step 3 of the algorithm **
## Deseasonalize the time series and estimate the trend parametrically.

## remove the seasonality
```{r}
deseasonalized <- temp - s.hat
```

## plot the deseasonalized time series.
```{r}
plot(year, deseasonalized, type="l",
     xlab="year", ylab="deseasonalized temperature")
```
## Part (f)
## Describe this plot.
This is the deseasonalized plot obtained after removing the seasonlaity and the deaseasonalized data contains only the trend estimate and noise. 

## We now fit a quadratic line model to the data.
```{r}
lm.deseasonalized <- lm(deseasonalized ~ year + I(year^2))
```

## Summarize the model obtained.
```{r}
summary(lm.deseasonalized)
```
## Plot the series and the estimated regression line.
```{r}
plot(year, deseasonalized, type="l",
     xlab="year", ylab="deseasonalized temperature")
lines(year, fitted(lm.deseasonalized), lty=2)
```

## Part (g)
## First write down the statistical model that has been fit to the deseasonalized data.
## Next, write down the estimated trend model. Explain why can we not assess the statistical significance of the regression parameters from this model output.
dt = yt - ^st
dt = -2.545e+05 + 2.544e+02*year - 6.358e-02*I(year^2)
The statistical significance of the regression parameters can not be accessed because of the fact that the deseasonlized trend component is not IID noise.

## ** Step 4 of the algorithm **
## Calculate the residuals and look at the ACF.

## calculate the residuals
```{r}
resids <- resid(lm.deseasonalized)
```

## Plot the residual series with a horizontal line at y=0.
```{r}
par(mfrow=c(2, 2))
plot(year, resids, type="l", xlab="year", ylab="estimate of noise process")
abline(h=0, lty=2)
```
## plot the acf
```{r}
acf(resids, main="")
```
## carry out the Ljung-Box test
```{r}
Box.test(resids, 22, type="Ljung-Box")
```
## draw a Q-Q plot of the residuals.
```{r}
qqnorm(resids, xlab="z-scores", ylab="estimate of noise process", main="")
qqline(resids)
```
## part (h)
## Using the figures, summarize the residuals.  In your answer make sure you respond to the following:
## - Do the residuals seem to be drawn from a stationary process?
Yes, the residuals seem to be drawn from a stationary process because the variance seems to be unfiorm and the mean is centered at zero and there are no abrupt changes in behaviour.
## - Do you believe that the residuals are a sample of IID noise?
Yes, from the plot of ACF, we can see that the residuals are a sample of IID noise because from the Box-Ljung test, p>alpha, the null hypothesis upholds and hence it is a IID noise. 
## - Do the residuals seem to be drawn from a normal distribution?
Yes, from the QQ plot, we can say that the residuals seem to be drawn from a normal distribution.

3.

```{r}
acf1 <- sapply(0:20, function(h) {
  sz <- 1
  sw <- 1
  phi <- 0.8
  if (h == 0) {
  1
  } else {
  ( sz^2*phi^h ) / (sz^2 + sw^2*(1-phi^2)^2)
  }
})
plot(acf1, type="h",x = 0:20, xlab = "Lag", ylab = "ACF")
```

```{r}
acf2 <- sapply(0:20, function(h) {
  sz <- 1
  sw <- 2
  phi <- 0.8
  if (h == 0) {
  1
  } else {
    ( sz^2*phi^h ) / (sz^2 + sw^2*(1-phi^2)^2)
  }
})
plot(acf2, type="h", x= 0:20, xlab = "Lag", ylab = "ACF")
```

From the plots of ACF for different lags, we can see that by increasing sigma_w^2 , the ACF decreases since the ACF and the value of sigma_w^2 are inversely proportional.